---
gettitle: "RR Peer Assignment 1"
author: "nvramamoorthy"
date: "16 May 2015"
output: html_document
---
#Reproducible Research : Peer Assignment 1

##Initial Setup


For the purpose of assignment a directory "~/Desktop/Reproducible Research/Peer-Assignment-1/RepData_PeerAssessment1" is created on my MacBook 
The suggested 

GitHub repository "forked from rdpeng/RepData_PeerAssessment1" needed is forked and cloned  on my DeskTop.

"activity.zip" was unziped and activity.csv file is kept for further processing

##Data

The data for this assignment was  downloaded from the course web site:

    Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

    steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

    date: The date on which the measurement was taken in YYYY-MM-DD format

    interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.



##Setting Working Directory

```{r}
 setwd("~/Desktop/Reproducible Research/Peer-Assignment-1/RepData_PeerAssessment1")
```


##Loading and preprocessing the data

```{r}
activity<-read.csv("activity.csv" , header=TRUE)
activity$date <- as.Date(activity$date) 

```
##Checking the contents of the Data
```{r}
dim(activity)
head(activity)
```






```{r, echo=FALSE}



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


##What is mean total number of steps taken per day?

For this part of the assignment,  the missing values were ignored in the dataset.

    Make a histogram of the total number of steps taken each day

    Calculate and report the mean and median total number of steps taken per day

```{r}
activity.ignore.na <- na.omit(activity) 

# sum steps by date
daily.steps <- rowsum(activity.ignore.na$steps, format(activity.ignore.na$date, '%Y-%m-%d')) 
daily.steps <- data.frame(daily.steps) 
names(daily.steps) <- ("steps") 

```

Plot histogram of the total number of steps taken each day:

```{r}
hist(daily.steps$steps, 
     main=" ",
     breaks=10,
     xlab="Total Number of Steps Taken Daily")
```
Compute  mean and median of steps:

```{r}
mean(daily.steps$steps);
median(daily.steps$steps)
```


##What is the average daily activity pattern?

    Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

    Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

    Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

    Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

    Create a new dataset that is equal to the original dataset but with the missing data filled in.

    Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

##Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

    Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

    Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:



